---
title: 'JHU Coursera: Statistical Inference Assignment1_2: '
author: "Dheeraj Agarwal"
date: "October 24, 2015"
output:
  pdf_document:
    toc: yes
  html_document:
    keep_md: yes
---

# SCOPE:
1. Load the ToothGrowth data and perform some basic exploratory data analyses.  
2. Provide a basic summary of the data.  
3. Use confidence intervals and/or hypothesis tests to compare tooth growth by supp and dose. (Only use the techniques from class, even if there's other approaches worth considering).  
4. State your conclusions and the assumptions needed for your conclusion  

***  
#### *Global Settings*  
The below global setting allow the code to be visible and switches the scientific notation of large numbers, if any, off.  
```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE, options(scipen=999))
library(ggplot2)
```  
***  
## **PART 1: DATA LOAD & EXPLORATORY ANALYIS**
______________________________________________________  

The below section will load the data as well as perform some function based basic exploratoy analysis. 
```{r explo_analysis1}
library(datasets)
data("ToothGrowth")
head(ToothGrowth, 5) # Display the first 5 rows and all columns
dim(ToothGrowth) # No of Rows and Columns in the entire data set
```  

All the three columns has the following classes: 
**`r class(ToothGrowth$len)`**, **`r class(ToothGrowth$supp)`** & **`r class(ToothGrowth$dose)`** respectively.

```{r explo_analysis2}
range(ToothGrowth$len)
unique (ToothGrowth$supp)
unique (ToothGrowth$dose)
group <- aggregate(ToothGrowth[,1], as.list(ToothGrowth[,2:3]), FUN=median)
colnames(group)[3] <- "MedianLength"
group
``` 

The above median calculations are also depicted below with the help of the box plots. 

```{r plots}
# Plotting the data for each quantity of dose and
 # delivery method against length of tooth in the guinea pigs.
ggplot(data = ToothGrowth, aes(x=as.factor(dose), y=len))+
        geom_boxplot(
                col = "black", 
                fill = c("indianred2", "dodgerblue", "green2"), 
                alpha = 0.4
        ) +
        labs(title = "Plot ", x = "Values", y = "TBD") + 
        facet_grid(supp ~ .)
```  

Based on the above boxplot, we can see that there is a marked shift in the tooth length as dosage increase. Based on the median values it also appears that that tooth length vary by the supplement type (OJ vs VC).  

## **PART 2: HYPOTHESIS FORMULATION & INFERENTIAL ANALYSIS**
______________________________________________________  

Given the above exploratory analysis, the:  
* First NULL Hypothesis or H0(1) is: OJ and VC have different effects on tooth length  
* Second NULL Hypothesis or H0(2) is: Dosage has an increaing effect on tooth length

**Data Prep**
```{r data_prep4_ht}
supp_val <- split(ToothGrowth$len, ToothGrowth$supp)
sapply(supp_val,var)
dose_val <- split(ToothGrowth$len, ToothGrowth$dose)
sapply(dose_val,var)
```

These variances are not close to each other hence any subsequent t-tests cannot be assumed to have a constant variance. Also the values cannot be considered paired for t-tests. This is because the tests were performed on different populations. That is the same subjects were not provided both the tests. Now, the reason a t-test is considered and not any other is because the t-test will help support (or contradict) the inferences made by the plots alone. That is test the hypothesis.

**Testing NULL Hypothesis 1**
```{r H0(1)}
t.test(
        ToothGrowth$len[ToothGrowth$supp == "VC"],
        ToothGrowth$len[ToothGrowth$supp == "OJ"], 
        paired=FALSE, var.equal=FALSE
)
```

From the values above, there are two important observations:
1. the **p value** is **0.06**
2. the 95% confidence interval is **-7.57 to 0.17**

That is the 95% confidence interval contains the value 0 and pvalue is greater than the usual standard of 5%. Hence we ***can reject the NULL***. It cannot be conclusively said if the supplment type makes an impactful difference.

**Testing NULL Hypothesis 2**

For this test, we will run t-tests for two samples at a time. Since we have 3 different dosages, we will have 3 samples with two dosages selected at a time for tests.

At first, comparing the 0.5 and 1 mg dosages
```{r H0(2)_1}
t.test(
        ToothGrowth$len[ToothGrowth$dose==1],
        ToothGrowth$len[ToothGrowth$dose==0.5], 
        paired = FALSE, var.equal = FALSE
)
```

Comparing the 0.5 and 2 mg dosage
```{r H0(2)_2}
t.test(
        ToothGrowth$len[ToothGrowth$dose==2],
        ToothGrowth$len[ToothGrowth$dose==0.5], 
        paired = FALSE, var.equal = FALSE
)
```

And finally comparing the 1 mg and 2 mg dosage
```{r H0(2)_3}
t.test(
        ToothGrowth$len[ToothGrowth$dose==2],
        ToothGrowth$len[ToothGrowth$dose==1], 
        paired = FALSE, var.equal = FALSE
)
```

Consolidating all the above results in a table. The confidence interval range is rounded to 2 decimals.

**Measure**     | **p-value**  | **95% confidence interval**
--------------- | ------------ | --------------
*0.5 & 1.0 mg*  |       0      | 06.28 - 11.98
*0.5 & 2.0 mg*  |       0      | 12.83 - 18.16
*1.0 & 2.0 mg*  |       0      | 03.77 - 9.00

In all cases, the **p-value** is **0**. Also the 95% confidence interval in all instances, do not contain **0** hence we ***cannot reject the NULL***. There is a marked impact of dosage on the tooth length.

***  
DISCLAIMER: This concludes my report based on my understanding of the problem statements. This report was created as part of Coursera Data Science Specialization::Reproducible Research course starting Oct 5th 2015. Please feel free to use it as a reference, however if you are taking the course yourself, please do not copy it.